---
title: "Integrated Assesment"
author: "S.Harripersad"
date: "09/02/2022"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
### Evaluation
If we can reach 95% accuracy at predicting whether or not a patient has heart disease during the proof of concept, we'll pursue the project.

# Predicting Heart disease using machine learning

This notebook looks into using various Python-based machine-learning and data science libraries in attempt to build a machine learning model capable of predicting whether or not someone has heart disease based on their medical attribute.

I am going to take the following approach:
1. Problem Definition.
2. Data.
3. Evaluation.
4. Features.
5. Modelling.
6. Experimentation

## Problem Definition
 
In a statement:
Given clinical parameters about a patient, can we predict whether our not they have heart disease. 

## Data

This heart disease dataset is curated by combining 5 popular heart disease datasets already available independently but not combined before. In this dataset, 5 heart datasets are combined over 11 common features which makes it the largest heart disease dataset available so far for research purposes. The five datasets used for its curation are:

*  Cleveland
*  Hungarian
*  Switzerland
*  Long Beach VA
*  Statlog (Heart) Data Set.

The original data came from: https://ieee-dataport.org/open-access/heart-disease-dataset-comprehensive

### Data Description - Features
##Create Data Dictionary

1.  age - Age, age in years(data type - numeric)
2.  sex - Sex(data type - binary):
    *   1 = male; 
    *   0 = female. 
3.  chest pain type - chest pain type(data type - nominal):
    *  Value: 1- Typical angina: chest pain related decrease blood supply to the heart;
    *  Value: 2 - Atypical angina: chest pain not related to heart;
    *  Value: 3 - Non-anginal pain: typically esophageal spasms (non heart related);
    *  Value: 4 - Asymptomatic: chest pain not showing signs of disease.
4.  resting blood pressure - resting bp s(in mm Hg on admission to the hospital), (data type numeric):
    *  anything above 130-140 is typically cause for concern.
5.  serum cholesterol - Cholesterol in mg/dl(data type numeric):
    *  serum = LDL + HDL + .2 * triglycerides;
    *  above 200 is cause for concern;
6.  fasting blood sugar - fasting blood sugar(1,0 > 120 mg/dl), (data type - binary): 
    *   1 = true; 
    *   0 = false; 
    *  '>126' mg/dL signals diabetes.
7.  resting electrocardiogram results -restecg results(data type - nominal):
    *   Value: 1 - Nothing to note;
    *   Value: 2 - ST-T Wave abnormality:
      *   can range from mild symptoms to severe problems;
      *   signals non-normal heart beat.
    *   Value: 3 - Possible or definite left ventricular hypertrophy:
      *   Enlarged heart's main pumping chamber.
8.  maximum heart rate achieved - max heart rate(71 - 202), (data type - numeric)
9.  exercise induced angina - exercise angina(data type binary):
    *   1 = yes; 
    *   0 = no.
10. oldpeak =ST - oldpeak(depression induced by exercise relative to rest), (data type - numeric):
    *   looks at stress of heart during exercise;
    *   unhealthy heart will stress more.
11. the slope of the peak exercise ST segment - ST slope(data type - nominal):
      *  Value: 1 - Up sloping: better heart rate with exercise (uncommon);
      *  Value: 2 - Flat sloping: minimal change (typical healthy heart);
      *  Value: 3 - Down sloping: signs of unhealthy heart.
12. target - have disease or not (1=yes, 0=no) (= the predicted attribute) (data type - binary)

Note: No personal identifiable information (PPI) can be found in the data set.

```{r}
# install the caret, rpart and ROCR packages from CRAN
if(require(caret) == FALSE){
  install.packages('caret', dependencies = TRUE)
  library(caret)
}
if(require(rpart) == FALSE){
  install.packages('rpart')
  library(rpart)
}
if(require(ROCR) == FALSE){
  install.packages('ROCR')
  library(ROCR)
}

# for a full list of models available in the caret packages:
#   http://topepo.github.io/caret/available-models.html
# for each model the list of tunable parameters is available
#   from the caret command modelLookup
```


# Loading package
```{r}
library(dplyr) #A Grammar of Data Manipulation
library(tibble) #modern take on data frames.
library(dlookr) #Tools for Data Diagnosis, Exploration, Transformation (main library)
library(tidyverse)
library(flextable)
library(Hmisc)
library(prettydoc)
library(corrr)
library(parsnip)
library(broom)
library(yardstick)
library(caret)
library(datarium)
library(ROCR)
library(forecast)
library(rms)
library(janitor)
library(vtree)
```

#### 2. Data exploration and preparation{#Data_exploration} 
```{r}
# read the data from a .csv file
heart <- read.csv("heart_statlog_cleveland_hungary_final.csv",na.strings = c("NA", ""))
```

```{r}
dim(heart)
```


```{r}
# inspect the data 
diagnose(heart) %>% flextable()
```
From the table above I can see that the Variables matches the data type as given in the data dictionary above.  For example age which is a variable matches with its' type which is an integer(numeric) and this is true for all variables and types. There are no missing values.  There is a potential problem  of having duplicate since this data set is a merger of four different heart disease data set.  There is also a need to check for ouliers.  


# Numeric value analysis

```{r}
diagnose_numeric(heart) %>% flextable()
```

Plotting for each feature.  
```{r}
plot_outlier(heart)
```

Upon analysis of this table I can see several problem:
1.  Variables with outliers include:
  *   sex;
  *   chest.pain.type
  *   resting.bp.s
  *   cholesterol
  *   fasting.blood.sugar;
  *   max heart.rate;
  *   ST.slop
  
2.  Missing values:
  *   cholesterol = 0.  This data is numeric and no one alive has a cholesterol reading 0 mg/dl.  It means that this most likely was missing data that was given a value of 0.

Lets us plot the outliers

#Graphical Analysis

```{r}
# generate a histogram for each variable (and show them on the same page)
#   note: titles and x labels are set to the name of the relevant variable
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,3))
hist(heart[, 1], main = names(heart)[1], xlab = names(heart)[1], xlim = c(0,100))
hist(heart[, 2], main = names(heart)[2], xlab = names(heart)[2], xlim = c(0,100))
hist(heart[, 3], main = names(heart)[3], xlab = names(heart)[3], xlim = c(0,100))
hist(heart[, 4], main = names(heart)[4], xlab = names(heart)[4], xlim = c(0,100))
hist(heart[, 5], main = names(heart)[5], xlab = names(heart)[5], xlim = c(0,100))
hist(heart[, 6], main = names(heart)[6], xlab = names(heart)[6], xlim = c(0,100))
hist(heart[, 7], main = names(heart)[7], xlab = names(heart)[7], xlim = c(0,100))
hist(heart[, 8], main = names(heart)[8], xlab = names(heart)[8], xlim = c(0,100))
hist(heart[, 9], main = names(heart)[9], xlab = names(heart)[9], xlim = c(0,100))
hist(heart[, 10], main = names(heart)[10], xlab = names(heart)[10], xlim = c(0,100))
hist(heart[, 11], main = names(heart)[11], xlab = names(heart)[11], xlim = c(0,100))
hist(heart[, 12], main = names(heart)[12], xlab = names(heart)[12], xlim = c(0,100))
par(opar)

# generate a density plot for each variable (and show them on the same page)
#   note: kernel density estimation may have tails outside the variable range!
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,3))
plot(density(heart[, 1]), main = names(heart)[1], xlab = names(heart)[1])
plot(density(heart[, 2]), main = names(heart)[2], xlab = names(heart)[2])
plot(density(heart[, 3]), main = names(heart)[3], xlab = names(heart)[3])
plot(density(heart[, 4]), main = names(heart)[4], xlab = names(heart)[4])
plot(density(heart[, 5]), main = names(heart)[5], xlab = names(heart)[5])
plot(density(heart[, 6]), main = names(heart)[6], xlab = names(heart)[6])
plot(density(heart[, 7]), main = names(heart)[7], xlab = names(heart)[7])
plot(density(heart[, 8]), main = names(heart)[8], xlab = names(heart)[8])
plot(density(heart[, 9]), main = names(heart)[9], xlab = names(heart)[9])
plot(density(heart[, 10]), main = names(heart)[10], xlab = names(heart)[10])
plot(density(heart[, 11]), main = names(heart)[11], xlab = names(heart)[11])
plot(density(heart[, 12]), main = names(heart)[12], xlab = names(heart)[12])
par(opar)

# generate a boxplot graph including horizontal background dashed lines
#   note: this can be done by
#     1. plotting an empty graph, i.e. with white boxes
#     2. adding the background dashed lines
#     3. plotting the coloured boxplot with the option add = T
opar <- par(no.readonly = TRUE)
boxplot(
  swiss,
  border = 'white',
  yaxt = 'n'
)
abline(h = seq(0,100,10), lty = 'dashed', lwd = 2, col = 'darkgrey')
boxplot(
  swiss,
  border = 'blue',
  yaxt = 'n',
  add = T
)
axis(2, seq(0,100,10))
par(opar)
```


Age - positive skewness, Kurtosis none 
Sex - negative skewness
Chest.pain.type - negative skewness
Resting.bp.s - negative skewness
Cholesterol - negative skewness
Fasting.blood - positive skewness
resting.ecg - positive skewness
max.heart.rate - positive skewness
exercise.agina - positive skewness
oldpeak - negative skewness
ST.slope - negative skewness
target - negative skewness

The plots confirms that the ouliers do exist in the seven features mentioend above.



```{r}
hist(heart$target)
```
Those suffering from heart disease is greater in numbers versus those that is not.  I wonder will a gender analysis show.  Looking at it below.  For men there is a greater percentage of men having the disease than women.  

```{r}
tabyl(heart, sex, target)
```
# Detecting missing/incorect data as identified in the variable cholesterol and identification of outliers.
After cleaning the data 

```{r}
#  Identifying which values == 0 in this column
#  heart$cholesterol == 0 will give all values = 0 as True and the rest as false.
#  Passing the above result to the column and replacing with NA
heart$cholesterol[heart$cholesterol == 0] <- NA
# Removing all instance of NA
heart_clean <- na.omit(heart)
# Check the size
dim(heart)
dim(heart_clean)
```


```{r}
boxplot(target ~ age, data = heart)
boxplot(target ~ sex, data = heart)
boxplot(target ~ chest.pain.type, data = heart)
boxplot(target ~ resting.bp.s, data = heart)
boxplot(target ~ cholesterol, data = heart)
boxplot(target ~ fasting.blood.sugar, data = heart)
boxplot(target ~ resting.ecg, data = heart)
boxplot(target ~ max.heart.rate, data = heart)
boxplot(target ~ exercise.angina, data = heart)
boxplot(target ~ oldpeak, data = heart)
boxplot(target ~ ST.slope, data = heart)
```
Boxplot analysis has confirm what numeric analysis table showed us and that is we have 7 variables with outliers.
 Variables with outliers include:
  *   sex;
  *   chest.pain.type
  *   resting.bp.s
  *   cholesterol
  *   fasting.blood.sugar;
  *   max heart.rate;
  *   ST.slop
  
I have chosen to remove outliers as identified by the boxplot analysis.

```{r}
# outliers was detected above using boxplot
#   note: the boxplot command returns a summary statistics object
#     this object can be assigned to a variable and inspected
heart_sex_boxplot <-boxplot(target ~ sex, data = heart)
heart_cpt_boxplot <-boxplot(target ~ chest.pain.type, data = heart)
heart_rbps_boxplot <-boxplot(target ~ resting.bp.s, data = heart)
heart_chol_boxplot <-boxplot(target ~ cholesterol, data = heart)
heart_fbs_boxplot <-boxplot(target ~ fasting.blood.sugar, data = heart)
heart_mhr_boxplot <-boxplot(target ~ max.heart.rate, data = heart)
heart_STs_boxplot <-boxplot(target ~ ST.slope, data = heart)
# inspect the boxplot statistics
#   note: the returened value is a list object
heart_sex_boxplot 
heart_cpt_boxplot 
heart_rbps_boxplot
heart_chol_boxplot
heart_fbs_boxplot 
heart_mhr_boxplot 
heart_STs_boxplot 
# outlier values can be extracted from this object
#   they are stored in the element 'out' of the list
heart_sex_boxplot$out 
heart_cpt_boxplot$out
heart_rbps_boxplot$out
heart_chol_boxplot$out
heart_fbs_boxplot$out
heart_mhr_boxplot$out
heart_STs_boxplot$out

# get the minimum value that is an outlier
min(heart_sex_boxplot$out) 
min(heart_cpt_boxplot$out)
min(heart_rbps_boxplot$out)
min(heart_chol_boxplot$out)
min(heart_fbs_boxplot$out)
min(heart_mhr_boxplot$out)
min(heart_STs_boxplot$out)

# outliers rows can be extracted by conditional selection
heart_clean[heart_clean$sex >= min(heart_sex_boxplot$out), ]
heart_clean[heart_clean$chest.pain.type >= min(heart_cpt_boxplot$out), ]
heart_clean[heart_clean$resting.bp.s >= min(heart_rbps_boxplot$out), ]
heart_clean[heart_clean$cholesterol >= min(heart_chol_boxplot$out), ]
heart_clean[heart_clean$fasting.blood.sugar >= min(heart_fbs_boxplot$out), ]
heart_clean[heart_clean$max.heart.rate >= min(heart_mhr_boxplot$out), ]
heart_clean[heart_clean$ST.slope >= min(heart_STs_boxplot$out), ]

```

```{r}
dim(heart)
dim(heart_clean)
```

```{r}
colnames(heart_clean)
```
I am going to finish this project in Google colab and the column names would be a problem "the name separated by ".".  I am going to rename the ones that is problematic.  

```{r}
colnames(heart_clean)[colnames(heart_clean) ==  "chest.pain.type" ] <- "chest_pain_type"
colnames(heart_clean)[colnames(heart_clean) == "resting.bp.s" ] <- "resting_bp_s"
colnames(heart_clean)[colnames(heart_clean) ==  "fasting.blood.sugar" ] <- "fasting_blood_sugar"
colnames(heart_clean)[colnames(heart_clean) ==  "resting.ecg" ] <- "resting_ecg"
colnames(heart_clean)[colnames(heart_clean) ==  "max.heart.rate" ] <- "max_heart_rate"
colnames(heart_clean)[colnames(heart_clean) ==  "exercise.angina" ] <- "exercise_angina" 
colnames(heart_clean)[colnames(heart_clean) ==  "ST.slope" ] <- "ST_slope"
```

```{r}
colnames(heart_clean)
```

```{r}
# perform PCA on the swiss dataset
#   note: variables are centered and scaled before analysis
pc_heart <- prcomp(heart_clean, center = T, scale. = T)

# inspect the attributes of the PCA object returned by prcomp
attributes(pc_heart)
# see value section of the help for the prcomp for more details
help(prcomp)
```

```{r}
# calculate the proportion of exaplained variance (PEV) from the std values
pc_heart_var <- pc_heart$sdev^2
pc_heart_var
pc_heart_PEV <- pc_heart_var / sum(pc_heart_var)
pc_heart_PEV

# plot the variance per PC
#   note: this can be done using the plot function on the prcomp object
plot(pc_heart)

# plot the cumulative value of PEV for increasing number of additional PCs
#   note: add an 80% threshold line to inform the feature extraction
#     according to the plot the first 3 PCs should be selected
opar <- par(no.readonly = TRUE)
plot(
  cumsum(pc_heart_PEV),
  ylim = c(0,1),
  xlab = 'PC',
  ylab = 'cumulative PEV',
  pch = 20,
  col = 'orange'
)
abline(h = 0.8, col = 'red', lty = 'dashed')
par(opar)

# get and inspect the loadings for each PC
#   note: loadings are reported as a rotation matrix (see lecture)
pc_heart_loadings <- pc_heart$rotation
pc_heart_loadings

# plot the loadings for the first three PCs as a barplot
#   note: two vectors for colours and labels are created for convenience
#     for details on the other parameters see the help for barplot and legend
opar <- par(no.readonly = TRUE)
colvector = c('red', 'orange', 'yellow', 'green', 'cyan', 'blue')
labvector = c('PC1', 'PC2', 'PC3')
barplot(
  pc_heart_loadings[,c(1:3)],
  beside = T,
  yaxt = 'n',
  names.arg = labvector,
  col = colvector,
  ylim = c(-1,1),
  border = 'white',
  ylab = 'loadings'
)
axis(2, seq(-1,1,0.1))
legend(
  'bottomright',
  bty = 'n',
  col = colvector,
  pch = 15,
  row.names(pc_heart_loadings)
)
par(opar)

# generate a biplot for each pair of important PCs (and show them on the same page)
#   note: the option choices is used to select the PCs - default is 1:2
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,2))
biplot(
  pc_heart,
  scale = TRUE,
  col = c('grey40','orange')
)
biplot(
  pc_heart,
  choices = c(1,3),
  scale = TRUE,
  col = c('grey40','orange')
)
biplot(
  pc_heart,
  choices = c(2,3),
  scale = TRUE,
  col = c('grey40','orange')
)
par(opar)
```


# Saving clean data to csv
```{r}
write.csv(heart_clean, "Clean_heart_data.csv")
```
